---
title: "Reproducible Research: Peer Assessment 1"
author: Albert de Roos
output: html_document
date: "`r Sys.Date()`"
keep_md: true
---
``` {r load_libraries, message=FALSE, results = "hide"}
library(dplyr)
Sys.setlocale("LC_TIME", "English")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/', echo=TRUE)
```

## Loading and preprocessing the data
The data file was unzipped using read.csv and the column with the dates was set to Date format. In order to get English-named weekdays, the locale on the machine was set to `r Sys.setlocale("LC_TIME", "English")`

```{r processing_data}
myrepdata <- read.csv("activity.csv")
myrepdata$date <- as.Date(myrepdata$date)
str(myrepdata)
```

## What is mean total number of steps taken per day?
In order to calculate the mean and median steps per day, first the data is grouped by the day and the sum per day is derived.

```{r total_steps_per_day}
stepsperday <- myrepdata %>% group_by(date) %>% summarize(sum = sum(steps))
str(stepsperday)
```

This yields a new data frame with `r nrow(stepsperday)` days consistent with the number of days in the 2 months of the data collection. The histogram of the steps taken per day and the mean and median are shown below. 

```{r histogram_total_steps_per_day}
hist(stepsperday$sum, breaks = 15, xlab = "total steps per day", main = "Steps per day distribution")
summary(stepsperday$sum)
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
